# Unidimensional categorical variables

In real-world datasets, categorical features are quite common but tricky during both the data pre-processing and visualization process. In this chapter, we will demonstrate several plotting options for the uni-dimensional categorical variables with ``ggplot``.

## Bar plot

There are two types of uni-dimensional categorical variables: nominal and ordinal. Here you will be shown how these variables should be plotted differently using bar plot under the same dataset.

### Nominal data

Nominal data is data with no fixed category order and should be sorted from highest to lowest count (left to right, or top to bottom)

```{r,fig.width=4.8, fig.height=3.6}
library(vcdExtra)
library(ggplot2)
library(forcats)
library(dplyr)

Accident %>%
  group_by(mode) %>%
  summarise(freq = sum(Freq)) %>%
  ggplot(aes(x=fct_reorder(mode,freq,.desc = TRUE),y=freq)) +
  geom_bar(stat = "identity",fill = "cornflowerblue") +
  ggtitle("Number of people with different modes in accident") +
  xlab("") +
  theme(panel.grid.major.x = element_blank())
```

… or top to bottom

```{r,fig.width=4.8, fig.height=3.6}
Accident %>%
  group_by(mode) %>%
  summarise(freq = sum(Freq)) %>%
  ggplot(aes(x=fct_rev(fct_reorder(mode,freq,.desc = TRUE)),y=freq)) +
  geom_bar(stat = "identity",fill = "cornflowerblue") +
  ggtitle("Number of people with different modes in accident") +
  coord_flip() +
  xlab("") +
  theme(panel.grid.major.x = element_blank())

```


### Ordinal data

Ordinal data is data having a fixed category order and need to sort it in logical order of the categories (left to right)

```{r,fig.width=4.8, fig.height=3.6}
Accident %>%
  group_by(age) %>%
  summarise(freq = sum(Freq)) %>%
  ggplot(aes(x=age,y=freq)) +
  geom_bar(stat = "identity",fill = "cornflowerblue") +
  ggtitle("Number of people of different ages in accident") +
  xlab("") +
  theme(panel.grid.major.x = element_blank())

```

Sort in logical order of the categories (starting at bottom OR top)

```{r,fig.width=4.8, fig.height=3.6}
Accident %>%
  group_by(age) %>%
  summarise(freq = sum(Freq)) %>%
  ggplot(aes(x=age,y=freq)) +
  geom_bar(stat = "identity",fill = "cornflowerblue") +
  ggtitle("Number of people of different ages in accident") +
  xlab("") +
  coord_flip() +
  theme(panel.grid.major.x = element_blank())

```

## Cleveland dot plot

```{r,fig.width=4.5, fig.height=6}
library(Lock5withR)
ggplot(USStates, aes(x = IQ, y = fct_reorder(State, IQ))) +
  geom_point(color = "blue") +
  ggtitle("Avg. IQ for US states") +
  ylab("") +
  theme_linedraw()

```

### Cleveland dot plot with multiple dots

Sort by Obese Rate

```{r,fig.width=4.5, fig.height=6}
library(tidyr)
USStates %>%
  select('State','Obese','HeavyDrinkers') %>%
  gather(key='type',value='percentage',Obese,HeavyDrinkers) %>%
  ggplot(aes(x=percentage, y=fct_reorder2(State,type=='Obese',percentage,.desc=FALSE), color = type)) +
  geom_point() +
  ggtitle("Obese rate & heavy drinker rate in US") +
  ylab("") +
  theme_linedraw()
```

### Cleveland dot plot with facets

```{r}
ggplot(USStates, aes(x = IQ, y = reorder(State, IQ))) +
  geom_point(color = "blue") +
  facet_grid(Pres2008 ~ ., scales = "free_y", space = "free_y") +
  ggtitle('IQ of US state residents facet by Pres2008') +
  xlab("IQ") +
  theme_linedraw() +
  theme(panel.grid.major.x = element_blank(), panel.grid.minor.x = element_blank())
```
